Specification and Verification of C3 linearization algorithm

### 2025-07-04 c3/article/article.tex
article almost written

\documentclass[runningheads]{llncs}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsmath,amssymb,todonotes,xspace}
\usepackage[hidelinks]{hyperref}
\usepackage{minted}
\usepackage[style=numeric-comp, sorting=none, backend=biber]{biblatex}
\addbibresource{refs.bib}

@@

\begin{abstract}
Multiple hierarchical inheritance is common among many programming languages, and
to produce a consistent and intuitive method resolution order (MRO) of the class hierarchy,
the C3 linearization algorithm is one of the most widely used.
However, the algorithm is not trivial, and because its widespread use,
there is a need for a formal specification and verification of this algorithm.
In this paper, we present a mathematical formal specification of the C3 algorithm,
 and its verification in OCaml code using Why3 and Cameleer.

@@

\subsection{Contributions}
To reach the goal of a verified implementation of the C3 linearization algorithm, we present the following contributions:
\begin{itemize}
    \item A mathematical formal specification of the C3 linearization algorithm, which includes the properties that the algorithm must satisfy.
    \item An OCaml implementation of the C3 linearization algorithm, which is verified using Why3 and Cameleer.
\end{itemize}

@@

\subsection{Cameleer and Why3}
Cameleer\autocite{pereiraCameleerDeductiveVerification2021} is an automated tool for deductive verification in OCaml. It uses comments written in GOSPEL\autocite{chargueraudGOSPELProvidingOCaml2019} (Generic OCaml Specification Language) to specify the properties of functions, and then leverages Why3\autocite{filliatreWhy3WherePrograms2013a} to verify these properties.\\
With GOSPEL, we can write pre-conditions, post-conditions, invariants, and variants using comments that start with a @ and are attached to a specific OCaml function. Take the OCaml function below as an example:
\inputminted[fontsize=\small,linenos,xleftmargin=20pt]{ocaml}{../../not_repeated/teste.ml}
In the function \texttt{filter\_heads} takes a list of lists as argument and returns a list of the heads of each sublist, the GOSPEL comments two post-conditions and a variant.
The first post-condition (lines 7 and 8) states that for every element \( y \) in the result list \( r \), exists a sublist \( x \) in the input list \( l \) such that the head of \( x \) is equal to \( y \).
The second post-condition (lines 9) states that for every sublist \( x \) in the input list \( l \), the head of \( x \) is in the result list \( r \).
And the variant (line 10) ensures the termination of the function by stating that the input list \( l \) is decreasing in size with each recursive call.\\
To run the verification, we can use the command \texttt{cameleer <file name>.ml}, which will open the Why3 IDE, like the one in Fig.\ref{fig:Why3FilterHeads}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{images/Why3FilterHead.png}
  \caption{\texttt{filter\_heads.ml} in Why3 IDE}
  \label{fig:Why3FilterHeads}
\end{figure}

@@

\subsubsection{Merge}
$\mer : \ordCC^* \Rightarrow \ordCC $

@@

>>>>

\documentclass[runningheads]{llncs}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsmath,amssymb,todonotes,xspace}
\usepackage{minted}
\usepackage{url}
\usepackage[hidelinks]{hyperref}           
\usepackage[style=numeric-comp, sorting=none, backend=biber, url=true]{biblatex}

\def\UrlBreaks{\do\/\do-}

\addbibresource{refs.bib}

@@

\begin{abstract}
Multiple inheritance is common in modern programming languages,
 and the C3 linearization algorithm is widely adopted to ensure a consistent and intuitive method resolution order (MRO). Despite its common use, particularly in Python, the algorithm is non-trivial, and rigorous formal specification and verification remain necessary.
In this paper, we present a mathematical formalization of the C3 algorithm and initiate its verification using OCaml, Why3, and Cameleer. Our work proves the termination and partial verification of the core components of the algorithm.
While complete verification of the linearization process is ongoing, our contributions represent an essential step toward a complete proof of C3's verification.

@@

\subsection{Contributions}
To reach the goal of a verified implementation of the C3 linearization algorithm, we present the following contributions:
\begin{itemize}
    \item A mathematical formal specification of the C3 linearization algorithm, which includes the properties that the algorithm must satisfy.
    \item An Ocaml implementation of the C3 linearization algorithm.
    \item A partial verification of the implementation using Why3 and Cameleer, demonstrating essential correctness properties and termination.
\end{itemize}

@@

\subsection{Cameleer and Why3}
Cameleer\autocite{pereiraCameleerDeductiveVerification2021} is an automated tool for deductive verification in OCaml. It uses comments written in GOSPEL\autocite{chargueraudGOSPELProvidingOCaml2019} (Generic OCaml Specification Language) to specify the properties of functions, and then leverages Why3\autocite{filliatreWhy3WherePrograms2013a} to verify these properties.\\
With GOSPEL, we can write pre-conditions, post-conditions, invariants, and variants using comments that start with a @ and are attached to a specific OCaml function. Take the OCaml function below as an example:
\inputminted[fontsize=\small,linenos,xleftmargin=20pt]{ocaml}{../../filter_heads/filter_heads.ml}
In the function \texttt{filter\_heads} takes a list of lists as argument and returns a list of the heads of each sublist, the GOSPEL comments two post-conditions and a variant.
The first post-condition (lines 7 and 8) states that for every element \( y \) in the result list \( r \), exists a sublist \( x \) in the input list \( l \) such that the head of \( x \) is equal to \( y \).
The second post-condition (lines 9) states that for every sublist \( x \) in the input list \( l \), the head of \( x \) is in the result list \( r \).
And the variant (line 10) ensures the termination of the function by stating that the input list \( l \) is decreasing in size with each recursive call.\\
To run the verification, we can use the command \texttt{cameleer <file name>.ml}, which will open the Why3 IDE, like the one in Fig.\ref{fig:Why3FilterHeads}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{images/Why3FilterHeads.png}
  \caption{\texttt{filter\_heads.ml} in Why3 IDE}
  \label{fig:Why3FilterHeads}
\end{figure}
As demonstrated by the green checkmark on the left side of the screen, the proof has been successfully verified. 
In the IDE, we proved these goals using provers, for our work we used three main provers,
 Alt-Ergo 2.6.0\autocite{sasAltErgoSMTSolver}, Z3 4.13.0\autocite{Z3ProverZ32025}, and CVC5 1.1.2\autocite{Cvc5} and for each goal, we can select which prover to use. To automate the proof Why3 provides a list of proof strategies including \texttt{Split\_vc}, \texttt{Auto\_level\_0}, \texttt{Auto\_level\_1}, 
\texttt{Auto\_level\_2}, and \texttt{Auto\_level\_3}. We also have commands to assist in the proof process, these commands are used to manipulate terms and hypotheses. All the commands and strategies are documented in Why3 documentation \autocite{12TechnicalInformations}.
Below is a list of commands that we utilized in our work:
\begin{itemize}
  \item \texttt{instantiate} 
  \item \texttt{induction\_arg\_ty\_lex}
  \item \texttt{intros} 
  \item \texttt{destruct}
  \item \texttt{destruct\_term}
  \item \texttt{unfold}
  \item \texttt{split\_vc}
\end{itemize}

@@

\subsubsection{Merge}
\label{sec:mergeFunctionMath}
$\mer : \ordCC^* \Rightarrow \ordCC $

@@

\section{Verification of C3 Linearization Algorithm}

In this project, we have two versions of the code, an original one and a version with some modifications to help the proof,
 and the code is divided into three parts: the \texttt{remove} where the remove of the head occurs, 
 the \texttt{merge} where a new list created follows the same order has the order lists and finally the \texttt{c3\_inearization} that will receive a class as argument and,
  in the version for the proof, a universe, a variable that contains all the class above in the hierarchy of that class.

\subsection{Module CLASS}

The code provided below represents a module that defines a class within the hierarchy:
\inputminted[firstline=4,lastline=22,fontsize=\small,linenos,xleftmargin=20pt]{ocaml}{../cameleer/c3.ml} 
The module was written to be as flexible as possible, so it could be easily implemented.
Later, it proved difficult to prove (as explained in Section~\ref{sec:proof-challenges}).

The function \texttt{get\_parents} takes a class and a list of lists as arguments and returns a list of parents of the class.
The only change from the original implementation is the list of lists as an argument in the function 
\texttt{get\_parents}, which was needed to ensure the hierarchy remained acyclic,
 passing as universe only the classes above in the hierarchy of the class.

\subsection{remove function}
Every time the function merge finds a head that has the three conditions (section~\ref{sec:mergeFunctionMath}) satisfied, the remove function is called.
This function removes a given head from all the list of lists, and it is defined as follows:
\inputminted[firstline=286,lastline=323,fontsize=\small,linenos,xleftmargin=20pt]{ocaml}{../cameleer/c3.ml}
The GOSPEL specification for \texttt{remove} includes preconditions and postconditions that aren't strictly needed to prove what \texttt{remove} itself does. But since \texttt{remove} is called inside \texttt{merge},
and \texttt{merge} relies on stronger guarantees, the extra conditions make it easier to prove \texttt{merge} correct.
So, even if removal could be verified with fewer assumptions, 
we added specific pre-conditions to simplify the verification of the functions that depend on it.\\
This function defines two major postconditions:
\begin{itemize}
  \item The argument \texttt{e} is not present in the any of the lists of the\\ argument \texttt{l} after the removal.
  \item The sum of the lengths of the lists of \texttt{l} are bigger than the\\ sum of the lengths of the lists of the result.
\end{itemize} 
The last postcondition is pivotal to prove the termination of the function \texttt{merge}. \\
To help the proof four lemmas were added to the code:
\begin{itemize}
  \item \texttt{list\_seq\_mem}: This lemma state that using both modules List or Sequence should produce the same result.
  \item \texttt{is\_removed\_not\_mem}: If an element is removed from the head (if the head is equal to the element) of a list and the tail does not contain that element then the element is not in the list.
  \item \texttt{is\_removed\_length\_for\_lists}: For all two lists of lists $l$ and $r$, and for each index $i$, the index $0$ is removed from $l[i]$ if $l[i][0] = e$ to produce $r[i]$, then the length of each $r[i]$ is less than or equal to the length of the corresponding $l[i]$.
  \item \texttt{is\_removed\_length}:If we remove the head of a list $l$ if the head is equal to $e$ and the head in $l$ is equal to $e$ then the length of $l$ decreases.
  \item \texttt{sum\_lengths\_of\_lists\_l\_e}: For all two lists of lists $l$ and $r$ of the same length, for each index $i$, the length of $l[i]$ is greater than or equal to the length of $r[i]$, then the sum of the lengths of $l$ is greater than or equal to the sum of the lengths of $r$.
  \item \texttt{sum\_lengths\_of\_lists\_l}: For all two lists of lists $l$ and $r$ of the same length, for each index $i$, the length of $l[i]$ is greater than or equal to the length of $r[i]$ and for one $j$, length of $l[j]$ is greater than $r[j]$, then the sum of the lengths of $l$ is greater than the sum of the lengths of $r$.
\end{itemize}
In the image (Fig. \ref{fig:Why3Remove}), we can see that the function was validated in the Why3 IDE.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{images/removeWhy3IDE.png}
  \caption{\texttt{remove} proof in Why3 IDE}
  \label{fig:Why3Remove} 
\end{figure}

\subsection{Detect order inconsistencies in a list of lists}
The function \texttt{merge} is the most complex function in the C3 linearization algorithm; this function receives a list of lists and returns a list where the order that each class has with each other respects in every list of the input.\\
To detect this, a precedence graph builds from the relations between the classes in the lists. For example, if we have the list of lists $((A, B, C),(C, D),(D, A))$, the graph in Fig.~\ref{fig:cyclicGraph} is built.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{images/cyclicGraph.png}
  \caption{Cyclic precedence graph}
  \label{fig:cyclicGraph}
\end{figure}
We can see that the graph is cyclic, and therefore, the order of the classes is inconsistent, and a linearization cannot be produced.
And it must exist an element that for every list of a list, this element is not in the tail of the list and it exists a head from a list that is equal to the element. % grammerly 
To detect the acyclicity of a graph, the following predicate was written:
\inputminted[firstline=64,lastline=80,fontsize=\small,linenos,xleftmargin=20pt]{ocaml}{../cameleer/c3.ml}
This predicate ensures that for the graph to be acyclic, no path exists in the graph where the first class is the same as the last class.
The lemmas created from this predicate are not proven in this project and assumed by postconditions and preconditions of the \texttt{merge} and \texttt{c3\_linearization} functions.
